Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-10-24 _tex/index.tex
Built site for gh-pages

Nevertheless, despite the growing number of studies on CJ, unsystematic
and fragmented research approaches in the literature have left several
critical issues unaddressed. These issues can be grouped into three main
categories: the disconnect between CJ's structural and measurement
model, the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in its measurement model, and the unclear
role and impact of comparison algorithms and the number of comparisons
on the method's reliability and validity. The following sections will
discuss each of these issues in detail,, followed by the introduction of
a theoretical model and its statistical translation, which aim to
address all three concerns simultaneously.

@@

In a typical CJ study, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model.
This model specifies how latent variables are estimated from observed
ones \citep{Hoyle_et_al_2023, Kline_et_al_2023}. In the context of CJ,
multiple judges engage in several rounds of pairwise comparisons to
assess the relative manifestation of a trait between two stimuli. Each
comparison generates a dichotomous outcome, indicating which stimulus is
perceived to exhibit a higher degree of the trait. The BTL model then
uses these observed outcomes to estimate scores that represent the
latent trait of interest
\citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.

Moreover, researchers often use these BTL-generated scores or their
transformations separately, applying additional analyses or hypothesis
testing. The CJ literature shows that these scores have been employed to
identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other scoring methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature warns against using estimated scores
for separate analyses and hypothesis testing. BTL-generated scores are
parameter estimates that inherently carry uncertainty. Neglecting this
uncertainty can inflate the precision and statistical power in
independent analyses and tests. This oversight, in turn, may increase
the risk of committing a type I error, where the null hypothesis is
incorrectly rejected \citep{McElreath_2020}. To mitigate this risk,
principles from Structural Equation Modeling (SEM)
\citep{Hoyle_et_al_2023, Kline_et_al_2023} recommend conducting such
analyses within the structural model of CJ rather than separately.
Therefore, an integrated approach combining both structural and
measurement models is advisable.

>>>>

Nevertheless, despite the growing number of studies on CJ, unsystematic
and fragmented research approaches in the literature have left several
critical issues unaddressed. This research primarily focuses on three:
the apparent disconnect between CJ's structural and measurement model,
the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
role and impact of comparison algorithms and the number of comparisons
on the method's reliability and validity. The following sections will
discuss each of these issues in detail, followed by the introduction of
a theoretical model and its statistical translation, which aim to
address all three concerns simultaneously.

@@

In a typical CJ study, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model.
This model specifies how latent variables are estimated from observed
variables \citep{Everitt_et_al_2010}. In CJ, multiple judges engage in
several rounds of pairwise comparisons to assess the relative
manifestation of a trait between two stimuli. Each comparison generates
a dichotomous outcome, indicating which stimulus is perceived to exhibit
a higher degree of the trait. The BTL model then uses these observed
outcomes to estimate scores that represent the latent trait of interest
\citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.

Moreover, researchers often use these BTL-generated scores or their
transformations separately, conducting further analyses or hypothesis
testing. The CJ literature highlights how these scores have been
employed to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other scoring methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature cautions against using estimated
scores for separate analyses and hypothesis testing. A key consideration
is that BTL-generated scores are parameter estimates that inherently
carry uncertainty. Ignoring this uncertainty when conducting separate
analyses and test can inflate their precision and statistical power.
This, in turn, increases the risk of committing a type I error, which is
when a null hypothesis is incorrectly rejected \citep{McElreath_2020}.
To mitigate this risk, principles from Structural Equation Modeling
(SEM) \citep{Hoyle_et_al_2023, Kline_et_al_2023} recommend conducting
these analyses within a structural model that accounts for both the
scores and their uncertainties, rather than treating them separately.
Thus, an integrated approach combining CJ's structural and measurement
models can offer significant advantages.
